Qualitative Probabilistic Programming

### 2015-05-28 paper/paper.tex
som stuff

\author{
Jessica Taylor \\
Department of Computer Science\\
Stanford University \\
Stanford, CA 94305
\texttt{jessica.liu.taylor@gmail.com}
\And
Noah Goodman \\
Department of ?? \\
Stanford University \\
Stanford, CA 94305
\texttt{...}
\And
}

@@

  \section{INTRODUCTION}

@@

  \section{SYNTAX}

  Quipp is implemented as a library for webppl programs.  Webppl [TODO cite] is a probabilistic programming language
  implemented in Javascript   Quipp programs are written as webppl programs that have access to special functions.

@@

  \begin{verbatim}
  dim = 2
  nclusters = 3
  PointType = Vector(dim, Double)
  ClusterType = Categorical(nclusters)

  get_point = rand_function(ClusterType, PointType)

  def sample():
    cluster = uniform_categorical(nclusters)
    return (cluster, get_point(cluster))
  \end{verbatim}

  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
  and one random function \texttt{get\_point}).
  Type annotations are necessary for random functions.  The type
  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
  and the type \texttt{Categorical(3)} expresses the fact that there are 3
  possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
  the distribution over clusters is uniform, but we do not know the distribution
  of points in each cluster.
  We will fill in the
  \texttt{get\_point} function with a learned function that will take a random
  sample from the given cluster.

  \section{FAMILY OF DISTRIBUTIONS}

  For unknown functions, the family of random functions used is a kind
    of generalized linear model.  We assume that the result
    is distributed from some exponential family whose natural
    parameters are determined from the arguments:

@@

    Here, $\eta$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
    and $g$ is the log partition function.

    To determine $\eta$ as a function of $x$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.  Features are the same as sufficient
    statistics for the categorical distribution, but while both $X$ and $X^2$ are
    sufficient statistics for the Gaussian distribution, only $X$ is a feature.

@@

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for entries not in the first
    row corresponding to sufficient statistics of $y$ that are not features.  This ensures that only the
    natural parameters for features are affected by $x$.

  \section{INFERENCE}

@@

  \section{EXAMPLES}

  \subsection{CLUSTERING}
{\small
\begin{verbatim}
dim = 2
nclusters = 3

PointType = Vector(dim, Double)
ClusterType = Categorical(nclusters)

get_cluster = rand_function(ClusterType)
get_point = rand_function(ClusterType, PointType)

def sample():
  cluster = get_cluster()
  return (cluster, get_point(cluster))
\end{verbatim}
}

@@

\subsection{NAIVE BAYES}

@@

  \subsection{FACTOR ANALYSIS}
{\small
\begin{verbatim}
num_components = 3
point_dim = 5

@@

(figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
  \subsection{HIDDEN MARKOV MODEL}
{\small
\begin{verbatim}
num_states = 5
num_observations = 3
sequence_length = 10

@@

\subsection{LATENT DIRICHLET ALLOCATION}
{\small
\begin{verbatim}
n_classes = 10
n_words = 100
n_words_per_document = 1000

@@

\subsection{NEURAL NETWORK}

@@

\subsection{A MORE COMPLEX MODEL}

@@

  \section{DISCUSSION}

@@

  In the future, it will be useful to support unbounded recursion in models (as in Church).  Furthermore, it will also be useful to expand the set of types supported.  It is possible to define reasonable default distributions for non-recursive algebraic data types, and it may also be possible to define them for recursive algebraic data types using catamorphisms.  Also, it will be useful to create a more usable interface to infer parameters and perform additional data pracessing given these parameters.

>>>>

\author{
Jessica Taylor \\
Department of Computer Science\\
Stanford University \\
Stanford, CA 94305
\texttt{jessica.liu.taylor@gmail.com}
\And
Andreas Stuhlm\"uller \\
Department of Brain and Cognitive Sciences \\
MIT \\
Cambridge, MA 02139 \\
\texttt{andreas@stuhlmueller.org}
\And
Noah Goodman \\
Department of Psychology \\
Stanford University \\
Stanford, CA 94305
\texttt{ngoodman@stanford.edu}
}

@@

  \section{Introduction}

@@

  \section{Syntax}

  Quipp is implemented as a library for webppl programs.  Webppl [TODO cite] is a probabilistic programming language
  implemented in Javascript. Quipp programs are written as webppl programs that have access to special functions.

@@

  \begin{verbatim}
var nclusters = 3;
var dim = 2;

var n = 100;

testParamInference(function(randFunction) {
  var pointType = Vector(dim, Double);
  var getPoint = randFunction(Categorical(nclusters), pointType);
  return function() {
    return repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      return observe('point' + i, getPoint, cluster);
    });
  };
});
  \end{verbatim}

  It is written as a generative model producing the observations.  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
  and one random function \texttt{getPoint}).
  Type annotations are necessary for random functions.  The type
  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
  and the type \texttt{Categorical(3)} expresses the fact that there are 3
  possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
  the distribution over clusters is uniform, but we do not know the distribution
  of points in each cluster.
  We will fill in the
  \texttt{getPoint} function with a learned function that will take a random
  sample from the given cluster.

  \section{Family of distributions}

  For unknown functions, the family of random functions used is a kind
    of generalized linear model.  We assume that the distribution
    of the function's return value
    is some exponential family whose natural
    parameters are determined from the arguments:

@@

    Here, $\eta(x)$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
    and $g$ is the log partition function.

    To determine $\eta(x)$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.  Features are the same as sufficient
    statistics for the categorical distribution, but while both $X$ and $X^2$ are
    sufficient statistics for the Gaussian distribution, only $X$ is a feature.

@@

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
    This ensures that only the
    natural parameters that are features of $y$ are affected by $x$.

  \section{Inference}

@@

  \section{Evaluation}

    To evaluate performance, for each model, we:
    \begin{itemize}
      \item
        Randomly generate parameters $\theta$
      \item
        Generate datasets $x_{train}, x_{test}$ using $\theta$
      \item
        Estimate $\log P(x_{test} | \theta)$
      \item
        Use the EM algorithm to infer approximate parameters $\hat{\theta}$ from $x_{train}$
      \item
        Estimate $\log P(x_{test} | \hat{\theta})$ and compare to $\log P(x_{test} | \theta)$
    \end{itemize}
    Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
    We use the Sequential Monte Carlo algorithm for this.  Between observations,

  \section{Examples}

  \subsection{Clustering}
{\small
\begin{verbatim}
var nclusters = 3;
var dim = 2;

var n = 100;

testParamInference(function(randFunction) {
  var pointType = Vector(dim, Double);
  var getPoint = randFunction(Categorical(nclusters), pointType);
  return function() {
    return repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      return observe('point' + i, getPoint, cluster);
    });
  };
});
\end{verbatim}
}

@@

\subsection{Naive Bayes}

@@

  \subsection{Factor analysis}
{\small
\begin{verbatim}
num_components = 3
point_dim = 5

@@

(figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
  \subsection{Hidden Markov model}
{\small
\begin{verbatim}
num_states = 5
num_observations = 3
sequence_length = 10

@@

\subsection{Latent Dirichlet allocation}
{\small
\begin{verbatim}
n_classes = 10
n_words = 100
n_words_per_document = 1000

@@

\subsection{Neural network}

@@

\subsection{A more complex model}

@@

  \section{Discussion}

@@

  In the future, it will be useful to expand the set of types supported.  It is possible to define reasonable default distributions for non-recursive algebraic data types, and it may also be possible to define them for recursive algebraic data types using catamorphisms.  Also, it will be useful to create a more usable interface to infer parameters and perform additional data pracessing given these parameters.
